Characterization of the microbial and fungal communities in ancient pottery from Neolithic southwestern China

### 2024-03-10 analysis/paper/paper.qmd
edit code chunk and add a plot

sig_genus_V4_pot <-
  composite_V4_sep_genus %>% # replace the region
  nest(data = -Genus) %>%
  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ pot, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
  unnest(test) %>%
  filter(p.value < 0.05)

V4_sig_family_all <- rbind(sig_family_V4_potex_con, sig_family_V4_pot)
V4_sig_family_name <- distinct(V4_sig_family_all, Family)

composite_V4_combine_pot_con <-
  composite_V4_sep_family %>%
  inner_join(V4_sig_family_name, by = "Family") %>%
  mutate(Family = str_remove(Family, "f__"),
         Family = str_replace(Family, "(.*)", "*\\1*")) %>%
  mutate(total_genera_sum = 100 * (total_genera_sum + 1/15000), # make it percentage and add a very small number
         pot = factor(group, levels = c("con", "pot-inside", "pot-outside", "soil"))) %>%
  ggplot(aes(x= total_genera_sum, y = Family, fill= group)) +
  geom_boxplot(aes(fill = group), color = "gray60", show.legend = TRUE) +
  scale_x_log10() +
  scale_fill_viridis_d(breaks = c("con", "pot-inside", "pot-outside", "soil"),
                       labels = c("control","pot-interior", "pot-exterior", "soil"),
                       direction = -1) +
  scale_y_discrete(limits=rev) +
  labs(x= "Relative abundance (%)", y= NULL) +
  theme_minimal() +
  theme(axis.text.y = element_markdown(size= 12),
        legend.title=element_text(size=12),
        legend.text=element_text(size=12))

ggsave(here::here("analysis","figures","V4_rel_sig_taxa_combine.png"),
       width = 9.5, height = 7, dpi = 360, units = "in")

@@

@@

V6_sig_family_all <- rbind(sig_family_V6_pot, sig_family_V6_potin_con, sig_family_V6_potex_con)
V6_sig_family_name <- distinct(V6_sig_family_all, Family)
V6_sig_genus_name <- sig_genus_V6_potin_con$Genus

composite_V6_combine_pot_con <-
  composite_V6_sep_genus %>%
  inner_join(sig_genus_V6_potin_con, by = "Genus") %>%
  mutate(Genus = str_remove(Genus, "g__"),
         Genus = str_replace(Genus, "(.*)", "*\\1*")) %>%
  mutate(total_genera_sum = 100 * (total_genera_sum + 1/15000), # make it percentage and add a very small number
         pot = factor(group, levels = c("con", "pot-inside", "pot-outside", "soil"))) %>%
  ggplot(aes(x= total_genera_sum, y = Genus, fill= group)) +
  geom_boxplot(aes(fill = group), color = "gray60", show.legend = TRUE) +
  scale_x_log10() +
  scale_fill_viridis_d(breaks = c("con", "pot-inside", "pot-outside", "soil"),
                       labels = c("control","pot-interior", "pot-exterior", "soil"),
                       direction = -1) +
  scale_y_discrete(limits=rev) +
  labs(x= "Relative abundance (%)", y= NULL) +
  theme_minimal() +
  theme(axis.text.y = element_markdown(size= 12),
        legend.title=element_text(size=12),
        legend.text=element_text(size=12))

```

The dominant classes include *Bacilli*, *Clostridia*, and *Negativicutes* within *Firmicutes*; *Actinobacteria*, *Rubrobacteria*, *Thermoleophilia* within *Actinobacteria*, *Deinococci* within *Deinococcus*, and *Gammaproteobacteria*, *Alphaproteobacteria*, and *Deltaproteobacteria* within *Proteobacteria*. At the family level, significant differences were observed between pottery exterior samples and control samples, and between pottery exterior and interior samples. *Paenibacillaceae*, *Erysipelotrichaceae*, and *Xiphinematobacteraceae* were more abundant in pottery exterior samples than those in control samples (P <0.05). In addition to *Paenibacillaceae* and *Xiphinematobacteraceae*, *Phycisphaeraceae* was also more abundant in pottery exterior samples than in pottery interior samples (P <0.05). At the genus level, *Candidatus_Xiphinematobacter* was more abundant in pottery exterior samples compared to pottery interior samples and control samples (P <0.01). Among the bacterial taxa showing significant differences, only *Paenibacillaceae* in the pottery exterior sample displayed relative abundances exceeding 10% on average, while others were below 1% (figure).

The V6 region has similar bacterial communities at the phylum level as the V4 region, with the exception of *Patescibacteria* (`r min(Pates$rel_ab)`- `r max(Pates$rel_ab)`%) in pottery samples, which was not detected in the reads of the V4 region (@fig-rel-ab: B). The dominant classes within *Patescibacteria* include *Parcubacteria*, *Microgenomatia*, and *Saccharimonadia*. We found a significant difference in the relative abundance of phyla between the pottery interior sample and soil samples from Gaoshan (P < 0.05), but none for Guijiabao samples. At the family level, *Paenibacillaceae* and *67_14* was more abundant in pottery exterior samples than in pottery interior samples (P <0.05). At the genus level, *Sphingomonas* was more abundant in the pottery interior samples than in the control samples (P <0.05).

@@

B: Bacterial communities at the family level differed significantly in relative abundance between pottery exterior and control samples, and between pottery exterior and interior samples

```{r}
#| label: fig-rel-ab
#| echo: false
#| cache: true
#| fig-cap: "A, B, C: The taxonomic composition of bacterial communities in pottery interior, pottery exterior, soil, and control samples based on sequences of the V4, V6, and V1 regions. D: The taxonomic composition of bacterial communities based on sequences of the ITS1 region."
library(here)
knitr::include_graphics(here("analysis", "figures", "all_rel_ab_combine.png"))
```

>>>>

sig_genus_V4_pot <-
  composite_V4_sep_genus %>% # replace the region
  nest(data = -Genus) %>%
  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ pot, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
  unnest(test) %>%
  filter(p.value < 0.05) %>% 
  filter(!Genus == "unclassified")

V4_sig_family_all <- rbind(sig_family_V4_potex_con, sig_family_V4_pot) %>% slice(1,2,5)

# join the original data of significance test 
composite_V4_genus_for_plot <-
  composite_V4_sep_genus %>%
  inner_join(sig_genus_V4_pot, by = "Genus") %>%
  mutate(Genus = str_remove(Genus, "g__")) %>% 
  rename(Taxon = Genus)

composite_V4_family_for_plot <-
  composite_V4_sep_family %>%
  inner_join(V4_sig_family_all, by = "Family") %>%
  mutate(Family = str_remove(Family, "f__")) %>% 
  rename(Taxon = Family)

# V4 plot
library(ggtext)
composite_V4_family_genus_com_plot <-
  rbind(composite_V4_genus_for_plot, composite_V4_family_for_plot) %>%
  mutate(Taxon = str_replace(Taxon, "(.*)", "*\\1*"),
         Taxon = str_replace(Taxon, "\\*Candidatus_(.*)\\*", "Candidatus<br>*\\1*")) %>%
  mutate(total_genera_sum = 100 * (total_genera_sum + 1/15000), # make it percentage and add a very small number
         pot = factor(group, levels = c("con", "pot-inside", "pot-outside", "soil"))) %>%
  ggplot(aes(x= total_genera_sum, y = Taxon, fill= group)) +
  geom_boxplot(aes(fill = group), color = "gray60", show.legend = TRUE) +
  scale_x_log10() +
  scale_fill_viridis_d(breaks = c("con", "pot-inside", "pot-outside", "soil"),
                       labels = c("control","pot-interior", "pot-exterior", "soil"), direction = -1) +
  scale_y_discrete(limits=rev) +
  labs(x= "Relative abundance (%)", y= NULL) +
  theme_minimal() +
  theme(axis.text.y = element_markdown(size= 12), legend.position = "none")

@@

sig_family_V6_potin_con <-
  composite_V6_sep_family %>% # replace the region
  nest(data = -Family) %>%
  mutate(test = map(.x= data, ~wilcox.test(total_genera_sum~ potin_con, data=.x, exact = FALSE) %>% tidy)) %>% #run broom first
  unnest(test) %>%
  filter(p.value < 0.05)

@@

V6_sig_family_all <- rbind(sig_family_V6_pot, sig_family_V6_potin_con) %>% slice(1:2)

# join the original data of significance test 
composite_V6_genus_for_plot <-
  composite_V6_sep_genus %>%
  inner_join(sig_genus_V6_potin_con, by = "Genus") %>%
  mutate(Genus = str_remove(Genus, "g__")) %>% 
  rename(Taxon = Genus)

composite_V6_family_for_plot <-
  composite_V6_sep_family %>%
  inner_join(V6_sig_family_all, by = "Family") %>%
  mutate(Family = str_remove(Family, "f__")) %>% 
  rename(Taxon = Family)

# V6 plot
composite_V6_family_genus_com_plot <-
  rbind(composite_V6_genus_for_plot, composite_V6_family_for_plot) %>%
  mutate(Taxon = str_replace(Taxon, "(.*)", "*\\1*")) %>%
  mutate(total_genera_sum = 100 * (total_genera_sum + 1/15000), # make it percentage and add a very small number
         pot = factor(group, levels = c("con", "pot-inside", "pot-outside", "soil"))) %>%
  ggplot(aes(x= total_genera_sum, y = Taxon, fill= group)) +
  geom_boxplot(aes(fill = group), color = "gray60", show.legend = TRUE) +
  scale_x_log10() +
  scale_fill_viridis_d(breaks = c("con", "pot-inside", "pot-outside", "soil"),
                       labels = c("control","pot-interior", "pot-exterior", "soil"),
                       direction = -1) +
  scale_y_discrete(limits=rev) +
  labs(x= "Relative abundance (%)", y= NULL) +
  theme_minimal() +
  theme(axis.text.y = element_markdown(size= 12),
        legend.title=element_text(size=12),
        legend.text=element_text(size=12))

library(cowplot)
V4_V6_rel_sig_taxa_combine <-
  plot_grid(composite_V4_family_genus_com_plot, composite_V6_family_genus_com_plot, rel_widths = c(0.8, 1), labels = "AUTO", ncol = 2)

ggsave(here::here("analysis","figures","V4_V6_rel_sig_taxa_combine.png"),
       width = 10, height = 4, dpi = 360, units = "in")
```

The dominant classes include *Bacilli*, *Clostridia*, and *Negativicutes* within *Firmicutes*; *Actinobacteria*, *Rubrobacteria*, *Thermoleophilia* within *Actinobacteria*, *Deinococci* within *Deinococcus*, and *Gammaproteobacteria*, *Alphaproteobacteria*, and *Deltaproteobacteria* within *Proteobacteria*. At the family level, significant differences were observed between pottery exterior samples and control samples, and between pottery exterior and interior samples. *Paenibacillaceae*, *Erysipelotrichaceae*, and *Xiphinematobacteraceae* were more abundant in pottery exterior samples than those in control samples (P <0.05; @fig-rel-ab-significant-plot: A). In addition to *Paenibacillaceae* and *Xiphinematobacteraceae*, *Phycisphaeraceae* was also more abundant in pottery exterior samples than in pottery interior samples (P <0.05). At the genus level, Candidatus *Xiphinematobacter* was more abundant in pottery exterior samples compared to pottery interior samples and control samples (P <0.01). Among the bacterial taxa showing significant differences, only *Paenibacillaceae* in the pottery exterior sample displayed relative abundances exceeding 10% on average, while others were below 1%.

The V6 region has similar bacterial communities at the phylum level as the V4 region, with the exception of *Patescibacteria* (`r min(Pates$rel_ab)`- `r max(Pates$rel_ab)`%) in pottery samples, which was not detected in the reads of the V4 region (@fig-rel-ab: B). The dominant classes within *Patescibacteria* include *Parcubacteria*, *Microgenomatia*, and *Saccharimonadia*. We found a significant difference in the relative abundance of phyla between the pottery interior sample and soil samples from Gaoshan (P < 0.05), but none for Guijiabao samples. At the family level, *Paenibacillaceae* and *67_14* was more abundant in pottery exterior samples than in pottery interior samples (P <0.05; @fig-rel-ab-significant-plot: B). At the genus level, *Sphingomonas* was more abundant in the pottery interior samples than in the control samples (P <0.05).

@@

```{r}
#| label: fig-rel-ab
#| echo: false
#| fig-cap: "A, B, C: The taxonomic composition of bacterial communities in pottery interior, pottery exterior, soil, and control samples based on sequences of the V4, V6, and V1 regions. D: The taxonomic composition of bacterial communities based on sequences of the ITS1 region."
library(here)
knitr::include_graphics(here("analysis", "figures", "all_rel_ab_combine.png"))
```

```{r}
#| label: fig-rel-ab-significant-plot
#| echo: false
#| fig-cap: "A: Bacterial communities at the family and genus level differed significantly in relative abundance between pottery exterior and control samples, and between pottery exterior and interior samples based on sequences of the V4 region. B: Bacterial communities at the family and genus level differed significantly in relative abundance between pottery exterior and interior samples, and between pottery interior and control samples based on sequences of the V6 region."
library(here)
knitr::include_graphics(here("analysis", "figures", "V4_V6_rel_sig_taxa_combine.png"))
```
